---
title: "Getting started with birtRcpp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with birtRcpp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`birtRcpp` fits joint models of response accuracy and response time. The main estimator is
`ecm()`: maximum marginal likelihood by the ECM algorithm (Meng & Rubin, 1993), with standard
errors, AIC and BIC, usually in about a second. `gibbs()` samples the posterior of the same
model, starting from the ECM solution. Both are written in C++. The package ports the Julia
package ExtendedRtIrtModeling.jl, with corrected full conditionals. Notation: j indexes
persons, i items.

| constructor | model |
|---|---|
| `mlirt()` | 2PL (or 1PL) with a latent regression of ability |
| `rtirt_null()` | 2PL + log-normal response times (van der Linden, 2007) |
| `rtirt_latreg()` | ... with latent regressions of ability and speed |
| `rtirt_latent()` | speed regressed on ability and covariates; `quantile =` for quantile regression |
| `rtirt_cross()` | cross-relations of ability with each log response time; `quantile =` |

## Data

`input_data()` takes responses, response times (in seconds) and centred covariates, either
as matrices or as column names of a data frame:

```{r}
library(birtRcpp)
set.seed(1)
cond <- set_cond(n_subj = 300, n_item = 8, n_feat = 2)
sim <- sim_data(cond, sim_para(cond, "latent"), "latent")
raw <- data.frame(sim$Y, sim$T, sim$X)
names(raw) <- c(paste0("y", 1:8), paste0("t", 1:8), "x1", "x2")
dat <- input_data(resp = paste0("y", 1:8), time = paste0("t", 1:8), cov = c("x1", "x2"), data = raw)
dat
```

## Maximum likelihood with ECM

ECM integrates ability over Gauss-Hermite nodes and speed in closed form. Each M step is
split into conditional maximizations, one per block of parameters, and each has a closed
form: the logistic item block through the Polya-Gamma bound (Polson, Scott & Windle, 2013),
the response-time and structural blocks as normal models. The log-likelihood never
decreases; SQUAREM accelerates it, and the standard errors come from the observed
information.

```{r}
fit <- ecm(rtirt_latent(dat))
fit
estimates(fit, pars = "beta|b_ability")   # est, se, z, Wald 95% interval
head(scores(fit))                         # EAP and posterior SD of ability and speed
c(logLik = logLik(fit), AIC = AIC(fit), BIC = BIC(fit))
```

Nested models compare by a likelihood-ratio test; `rtirt_null()` is `rtirt_latreg()`
without covariates:

```{r}
f0 <- ecm(rtirt_null(dat))      # warns that rtirt_null() ignores the covariates
f1 <- ecm(rtirt_latreg(dat))
lr <- 2 * (logLik(f1) - logLik(f0))
c(LR = lr, df = f1$ecm$npar - f0$ecm$npar,
  p = pchisq(lr, f1$ecm$npar - f0$ecm$npar, lower.tail = FALSE))
```

Other settings: `nodes` (Gauss-Hermite nodes, default 41), `max_iter`, `tol`, `se`,
`accelerate`, and `prior = "map"` with `priors = rtirt_priors(...)` for the posterior mode
in the item parameters (small samples).

## Gibbs from the ECM solution

```{r}
g <- gibbs(fit, n_iter = 1500, n_chain = 2, init = "ecm", seed = 1, verbose = FALSE)
estimates(g, pars = "beta|b_ability")     # posterior mean, SD, 95% interval, R-hat, ESS
convergence(g)
```

## Quantile models (Gibbs only)

Quantile models use the asymmetric Laplace working likelihood, so `ecm()` refuses them.
For several quantile levels, give a vector; the result holds one model per level:

```{r}
qs <- gibbs(rtirt_latent(dat, quantile = c(0.25, 0.5, 0.75)), n_iter = 1000, n_chain = 2,
            seed = 1, verbose = FALSE)
qs
```

Use longer chains in practice (the defaults are `n_iter = 5000`, `n_chain = 4`) and check
`convergence()`. The tutorial `system.file("tutorial", "birtRcpp_intro.qmd", package = "birtRcpp")`
(in Traditional Chinese) explains every line and output.

## References

Meng, X.-L., & Rubin, D. B. (1993). Maximum likelihood estimation via the ECM algorithm: A
general framework. *Biometrika, 80*, 267-278.

Polson, N. G., Scott, J. G., & Windle, J. (2013). Bayesian inference for logistic models using
Polya-Gamma latent variables. *Journal of the American Statistical Association, 108*, 1339-1349.

van der Linden, W. J. (2007). A hierarchical framework for modeling speed and accuracy on test
items. *Psychometrika, 72*, 287-308.
